---
title: "Spectral Analysis of Traffic Accidents in New York's Capital Region"
subtitle: |
  |
  | by
  | Michael Barr
  |
  |
  | A Thesis Submitted to the University at Albany, State University of New York in Partial Fulfillment of the Requirements for the Degree of Master of Science, Biostatistics
  |
  |
  | College of Integrated Health Sciences
  | Department of Biostatistics
date: "May 2026"
documentclass: report
output: 
  pdf_document:
    fig_caption: yes
geometry: margin=1in
fontsize: 12pt
header-includes:
  - \usepackage[font={footnotesize}]{caption}
  - \setlength{\parindent}{2em}
  - \usepackage{indentfirst}
  - \usepackage{setspace}
  - \doublespacing
  - \setlength{\parskip}{0pt} 
  - \usepackage{etoolbox}
  - \AtBeginEnvironment{Shaded}{\singlespace}
  - \usepackage{endnotes}
  - \let\footnote=\endnote
  - \usepackage{hanging}
---

```{r opts, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      scipen=999,
                      digits = 6,
                      cache.path = "cache/")
```

\pagenumbering{roman}
\setcounter{page}{2}

\centerline{\Large\bf Abstract}

In this paper we estimate the spectral density of traffic accident events in the Capital Distict, NY area using a band-pass filter known as the Kolmogorov-Zurbenko Fourier Transform (KZFT). The source data is provided by Moosavi, et al. (2019) and originally captured from various public entities and sensors in the road network. KZFT and spectral analysis methods in general are more typically applied to signals which are highly determined by physical forces and processes - objects that spin, oscillate, vibrate, orbit, and so on. Spectral density estimation with KZFT suppresses noise to reveal a greater number of constituent frequencies in the signal compared to standard fourier analysis. We are able to produce more accurate estimated energies, in particular for the many lesser harmonics of the fundamental weekly frequency detected in accident patterns. Signal reconstruction based on KZFT produces the approximate weekly pattern of this noisy signal which is not governed by any underlying physical processes transmitting sinusoidal waveform energy. Our novel application of these methods to population patterns measured by highly random loss events demonstrates their potential for greater adoption among researchers working with signals from varied fields.

```{r libs, include = FALSE}
library(lubridate)
library(dplyr)
library(mgcv)
library(parallel)
library(ggplot2)
library(gridExtra)
library(scales)
library(tidyr)
library(doParallel)
# library(extrafont)
source("kzft.R")

# font_import("/usr/local/lib/R/site-library/extrafontdb/metrics/")
# loadfonts(device = "pdf")
# theme_set(theme_bw(base_family = "CM Roman"))
```

```{r load_data, cache=TRUE, results = FALSE}
acc <- read.csv("US_Accidents_March23.csv")

cap <- acc %>%
  filter(Zipcode %in% c("12161", "12158", "12067", "12033", "12061", "12144",
                        "12054", "12077", "12202", "12209", "12208", "12207",
                        "12210", "12204", "12189", "12183", "12047", "12110",
                        "12205", "12222", "12226", "12206", "12159", "12151",
                        "12084", "12203", "12303", "12304", "12306", "12130",
                        "12065", "12309", "12307", "12155", "12132", "12136",
                        "12060", "12019", "12045", "12192")) %>%
  mutate(across(Start_Time:End_Time, ~ as.POSIXct(.x, tz = 'EST')))

n_events <- cap %>%
    filter(Start_Time >= as.POSIXct("2019-01-01 00:00:00", tz = 'EST'),
           Start_Time < as.POSIXct("2020-01-01 00:00:00", tz = 'EST')) %>%
  nrow()

dat <- tibble(t_start = seq(from = as.POSIXct("2016-03-24 00:00:00", tz = 'EST'),
                       to = as.POSIXct("2023-03-31 00:00:00", tz = 'EST'),
                       by = "hour"),
              t_end = t_start + 60^2) %>%
  mutate(t = row_number()) %>%
  relocate(t) %>%
  left_join(cap, join_by(overlaps(t_start, t_end, Start_Time, End_Time, bounds = "[)") )) %>%
  mutate(power = difftime(pmin(t_end, End_Time),
                          pmax(t_start, Start_Time), units = "hours") %>%
           as.numeric() %>%
           tidyr::replace_na(0)) %>%
  filter(t_start >= as.POSIXct("2019-01-01 00:00:00", tz = 'EST'),
         t_start < as.POSIXct("2020-01-01 00:00:00", tz = 'EST')) %>%
  group_by(t, t_start, t_end) %>%
  summarise(x = sum(power)) %>%
  ungroup()

rm(acc)
gc()
saveRDS(dat, "dat.RDS")
```

\newpage
\pagenumbering{arabic}

## Introduction

High frequency, remotely-sensed data is increasingly available to researchers in industry, government, and academia. Sensors have long been utilized to measure and monitor a wide variety of phenomena and conditions in environmental (air quality, ocean tides, atmospheric pressure, seismic activity), industrial or mechanical equipment (heat, vibration, alignment, etc. within motors, gearboxes, pumps, and various rotating equipment), built infrastructure (e.g. vibration sensors for structural health monitoring of bridges), and many other areas. Technological innovation on multiple fronts in recent years has reduced the size and cost of sensor hardware, increased options for connectivity (i.e. with short-range/bluetooth connectivity to smartphones and local wireless networks), and exponentially reduced the cost of physical data storage. These developments have fostered a massive expansion of "smart" product offerings for consumer and commercial applications including wearable biometric monitoring devices, home systems monitoring (moisture/leak and arc fault detection), and telematics devices which measure aspects of vehicle usage and driving behavior (acceleration/deceleration, speed, time of day). Many of these and other applications are linked to insurance products with the purpose of measuring, pricing for, and reducing the risk of loss events affecting health or property.

The increased deployment of sensors and the data streams they produce demands statistical methodologies suited for their analysis. A series of measurements taken sequentially over time - whether at regular or irregularly spaced intervals - is a time series. Methods for the analysis of time series have perhaps received less attention than warranted, possibly due to the concurrent development and focus on new machine learning techniques for regression analysis and advanced prediction systems in the context of data that is tabular, un-ordered, and rich with covariates. To the extent that methods for time series are given attention it is usually limited to time-domain methods - the auto-regressive model and its extensions.

The task of this paper is to demonstrate the relevance of "the other half" of time series analysis - spectral analysis - for data we may not immediately see as calling for it. Part I of this paper provides a brief motivation for spectral analysis with an overview of key concepts building to the frequency domain and the (Fast) Fourier Transform. From this groundwork we proceed to discuss a class of methods known as Kolmogorov-Zurbenko (KZ) Filters. In Part II we discuss in detail the extension of the KZ filter to spectral density analysis with the Kolmogorov-Zurbenko Fourier Transform (KZFT) band-pass filter. Part III demonstrates application of KZFT to a traffic accident event data set sourced from publicly available, streaming APIs which monitor and report traffic incidents in real-time.

## Part I: Motivation

### Temporal and Frequency Domains

```{r, echo = FALSE, cache = TRUE}
set.seed(1234)
simdat <- tibble(time = 0:(24*360),
                 f1 = ((time)/(12.5)) %% 1, 
                 f2 = ((time)/(10)) %% 1, 
                 f3 = ((time)/((24*7))) %% 1, 
                 c1 = 1.5 * sin(2 * pi * 1/(12.5) * time + 0),
                 c2 = .5 * sin(2 * pi * 1/(10) * time + pi/2),
                 c3 = 1 * sin(2 * pi * 1/(24*7) * time + pi/3),
                 y = c1 + c2 + c3,
                 r = rnorm(24*360+1, 0, 2),
                 x = y + r,
                 r2 = r*3,
                 x2 = y + r2)

sim_snr <- var(simdat$y)/var(simdat$r)
sim_snr2 <- var(simdat$y)/var(simdat$r2)
```

Assume a random variable $X_t$ occurring in time which is the result of some data generating process involving deterministic and stochastic components. A time-series is a sequence of measurements $\{X(t): t\in 0,1,2,...,n\}$ taken of $X_t$, recorded successively, indexed and ordered in time. $X_t$ may be multivariate - each measurement $X(t)$ a vector of values - but for simplicity we will only deal with the univariate case - we assume $X_t$ is scalar-valued. We will also assume $X_t$ is real-valued though this is not strictly necessary. A time series is the result of a data generating process as well as some repeatable means of obtaining and recording measurements - which may be imprecise - of the observable value at points in time. We will refer to the means of obtaining measurements as a *sensor*. A series of measurements taken of $X_t$ is often referred to as the "raw" signal or simply the *signal*.

Methods for time-series analysis may seek to describe the deterministic and stochastic components embedded in $X_t$, often by estimating a model from the data. Analysis methods can be classed into two broad categories. *Time-domain* methods work with the readings or *total energy* of the observed signal in the time domain we naturally perceive it. This includes any approach which views the data temporally - whether calculating a simple moving average, estimating an AR(p) model and its variations, or training an LSTM. Such methods are well-suited for signals which do not contain many underlying cyclical components - e.g. when $X_t$ is highly determined by recent lagged values $X_{t-1}, X_{t-2},...$ and so on. A brownian motion in its discrete form, the random walk, is the archetypical example.

When the data generating process underlying a signal is more complex - that is, when it is governed by multiple cyclical components - then analysis in the time domain quickly becomes intractable. *Frequency-domain* methods begin with some means of transforming a time-indexed signal into the frequency domain in order that we may better detect and describe the cyclical components ("constituents" or "waveforms") embedded in a complex and potentially noisy signal.

Figure 1 illustrates the issue with a simulated signal (top) generated by three cyclical constituents (together the "systematic component") and a white noise (the stochastic component). The expected value of the generating function is shown in total (second panel) and decomposed into its three constituents (third). The stochastic component is also shown (bottom).

```{r simdat, fig.show = 'hold', fig.height = 3.6, fig.cap = "$X(t) = 1.5 \\sin(\\frac{2 \\pi}{12.5}) + .5 \\sin(\\frac{2 \\pi}{10}) + \\sin(\\frac{2 \\pi}{168}) + N(0, \\sigma = 2)$"}
grid.arrange(
  nrow = 4,
  heights = c(.23,.23,.23, .31),
  simdat %>%
    filter(time < 240) %>%
    ggplot() +
    geom_line(aes(time, x)) +
    scale_y_continuous(name = "Signal") +
    theme(axis.title.y = element_text(size = 8),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()),
  
  simdat %>%
    filter(time < 240) %>%
    ggplot() +
    geom_line(aes(time, y)) +
    scale_y_continuous(limits = c(-3,3),
                       name = "Deterministic") +
    theme(axis.title.y = element_text(size = 8),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()),
  
  simdat %>%
    filter(time < 240) %>%
    select(time, c1:c3) %>%
    pivot_longer(cols = c(c1,c2,c3), names_to = "constituent", values_to = "energy") %>%
    ggplot(aes(time, energy, color = constituent)) +
    geom_line() +
    scale_y_continuous(limits = c(-3,3),
                       name = "Constituents") +
    theme(legend.position = "none",
          axis.title.y = element_text(size = 8),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()),
  
  simdat %>%
    filter(time < 240) %>%
    ggplot() +
    geom_line(aes(time, r)) +
    theme(axis.title.y = element_text(size = 8),
          axis.title.x = element_text(size = 8)) +
    scale_y_continuous(name = "White Noise")
)
```

Even a relatively simple signal like above poses a challenge for analysis in the time domain. Consider the de-noised signal - after removing the stochastic component it remains not obvious what underlies the pattern we observe in time. We can detect the dominant frequency at .08 (period = 12.5), but the relative magnitude at this frequency varies as it is modulated by phase-alignment with the waveform at frequency .10 every 4th period. The absolute magnitude is also seen to cycle more slowly with the third constituent. None of this is apparent in the raw signal, however, which appears very messy in the time domain despite a relatively high signal-to-noise ratio of .446. Though an AR(2) model can be used to perfectly describe a single sine wave without noise, the order required becomes prohibitively high as a signal becomes more complex. Analyzing partial autocorrelation is not a help because - like the AR(p) model - it is based in the assumption that the greatest predictive information is found in the nearest lagged values. In fact the predictive information for the next value $X(t)$ is found in the lags corresponding to the cycle times (multiples of$\frac{1}{f_{\nu}}$) of each waveform embedded in the signal - e.g. $X(t-10)$, $X(t-20)$,...,$X(t-\frac{c}{0.1})$ for a waveform present at frequency $f_{.1}$.

```{r AR, fig.show='hold', out.width = '50%', fig.cap = "(left) Using PACF to detect the order, or key lags, of an AR(p) model for a single sine wave at $f_{\\frac{1}{12}}$ without noise; (right) PACF applied to the simulated $X_t$ with SNR=.446"}
par(mar=c(4,4,.1,.1))
pacf(simdat$c1, lag.max = 15, main = NA)
pacf(simdat$x, lag.max = 200, main = NA)
```

This signal can be neatly summarized within the frequency domain. A periodogram represents an actual or, more often, the estimated power spectrum, that is, the power ($=\text{energy}^2$) of assumed sinusoid waveforms oscillating at each observable "frequency bin" from $f_0$ (the "DC component" or mean power level) up to the Nyquist limit $\frac{f_S}{2}$ (half the sampling rate of the signal). The power spectrum behind a signal may in some cases be known but usually, for a signal found "in nature" and measured in a research context, it must be estimated in the presence of noise. If we can identify the constituent frequencies which determine a signal, estimate their respective energy levels and phase or offset at $t=0$, then we can describe and reconstruct the non-random component of the signal. We can also use the information to forecast future values of $X_t$ yet to be realized, assuming the process is stationary. Figure 3 shows the (known) PSD of the deterministic component generating the above signal.

```{r sim_period, fig.align='center', out.width='50%', fig.cap="Theoretical periodogram (systematic component) of simulated signal"}
ggplot() +
  geom_segment(aes(x = 1/10, xend = 1/10, y = 0, yend = .5^2)) +
  geom_segment(aes(x = 1/12.5, xend = 1/12.5, y = 0, yend = 1.5^2)) +
  geom_segment(aes(x = 1/(24*7), xend = 1/(24*7), y = 0, yend = 1^2)) +
  geom_segment(aes(x = 0, xend = .5, y = 0)) +
  scale_x_continuous(limits = c(0,.5),
                     name = "Frequency") +
  labs(y = expression(Power==Energy^2)) +
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank()) 
```

The basic task of spectral analysis is to decompose a potentially complex and noisy signal into a set of non-random constituent waveforms and residual variance or unexplained noise by estimating the distribution of power across a spectrum of observable frequencies, also known as the Power Spectral Density (PSD). The periodogram is classically defined as $\frac{1}{N}|FFT(X_t)|^2$ based on the Fast Fourier Transform of $X_t$. As an estimator for the PSD this approach is asymptotically unbiased but also inconsistent and high variance - the result is often a highly noisy power spectrum which does not become smoother with greater samples. We desire an estimator $\theta$ which is not necessarily unbiased but instead optimizes a trade-off between bias and variance to minimize mean square error of the estimator.

$$MSE(\theta) = bias_{\theta}^2 + var(\theta)$$
We now turn to reviewing the foundational method of spectral density analysis known as the fourier transform.

### Fourier Transform

It is well known that any function $X(t)$ defined over a self-symmetric domain can be decomposed into even and odd parts, and therefore uniquely expressed as the sum of an even and odd function.
$$
X(t) = X_e(t) + X_o(t)
$$
\noindent
$cos(t)$ is an even function - i.e. it is symmetrical about the y-axis ($cos(-x) = cos(x)$), and $sin(t)$ is an odd function ($sin(-t) = -sin(t)$). 

Of course a sine wave is just a cosine with phase shift $\phi=\frac{\pi}{2}$. In general, a sine wave with phase offset can be expressed as a linear combination of a sine and cosine each without phase:  $a\sin(t + \phi) = a \cos(\phi) \sin(t) + a \sin(\phi) \cos(t)$. Put another way, one sine and one cosine function with identical period together form a complete (and othogonal) basis set for expressing an arbitrary sine function of the same period. This fact can be generalized to the complex case:
$$ia\sin(t + \phi) = z \big(\cos(t) + i\sin(t)\big)$$
\noindent
where z is a complex-valued constant containing both $a$ and $\phi$. We can allow that the imaginary part of the left term may be 0 - so we can express a sinusoid on the real plain with non-zero phase as the scaled sum of complex-valued sine and cosine waves without phase.

Finally, by Euler's identity we know that $e^{jt} = \cos(t) + j\sin(t)$, so the expression above can be restated as
$$ja\sin(t + \phi) = ze^{jt}$$

The Fourier Transform allows us to express any function $f(t)$ over a time domain as the integral of sine functions $F(\nu)$ spanning the frequency domain, each of them parameterized by some amplitude and phase.
$$
\begin{aligned}
f(t) &= \frac{1}{2\pi} \int^{\infty}_{-\infty} F(\nu)e^{j\nu (t)}d\nu \\
F(\nu) &= A(\nu)e^{j\phi(\nu)}
\end{aligned}
$$
Each $F(\nu)$ is a complex exponential at a given frequency. The set of all $F(\nu)$ forms a complete and orthogonal basis in the frequency dimension - they are pairwise independent and can be linearly combined to represent an arbitrary function.

Though we experience it every day, typically we cannot work with data in continuous time. Either our sensors will have a sampling rate or an analogue (continuous) signal must be digitized (sampled) for storage. So all time-series will be defined by a sampling rate and our methods will work in discrete time. The Discrete Fourier Transform (DFT) analogue to the above is more relevant in practice for regularly spaced signal data. The DFT is defined as
$$
X_k = \sum_{n=0}^{N-1}x_n e^{-j2\pi k \frac{n}{N}}
$$
\noindent
with $k$ the frequency index, $n$ the time index, and $N$ the number of samples in the series. The frequency index is the period for each waveform - a cycle repeating every 24 hours has frequency 1/24 = .0416667 in a signal with hourly sampling rate, and frequency 1/288 = .0034722 given readings every 5 minutes. In the DFT output these will be located at index 24 (the 25th frequency bin) or 288, the input signal and transformed output each indexed from 0.

Spectral density analysis was enabled in principle by the development of what came to be known as the Fourier Transform, based on work published by Jean-Baptiste Joseph Fourier in the 1820s in his study of heat transfer (Valentinuzzi, 2016).^[Fourier first wrote down his equations in an 1807 manuscript, and later in his 1822 published work. But as is often the case it appears Carl Friedrich Gauss was first to discover the method in 1805 along with an efficient "divide and conquer" algorithm for its computation which closely resembles the FFT algorithm developed 160 years later.] But it was its implementation via the Cooley-Tukey algorithm - the Fast Fourier Transform (FFT) - in 1965 that made spectral analysis feasible in the modern setting. Most algorithms today are implemented using FFT rather than direct calculation from the definition due to the difference in computational complexity. Direct calculation of the DFT from its definition requires $O(N^2)$ operations for the full spectrum, while FFT reduces this to $O(N\log N)$. This means for a series of 1,000 samples the DFT performs 1,000,000 operations while FFT requires only 6,908. The performance gain comes with a key limitation - missing values cannot be present in the input signal due to the recursive approach of the FFT algorithm.

While the FFT-based periodogram is a useful tool for many cases, it does not have the properties needed to reliably serve as an estimator for the power spectrum, in particular as signals become more complex and when signal-to-noise is low. More commonly approaches to PSD estimation will build from FFT a consistent and/or lower variance estimator either by smoothing the above result or through strategies of repeated application on subsets of the total signal. Variations include application of spline smoothers to the spectral density, different strategies fro breaking up the original signal into blocks and applying FFT to each separately before combining (averaging) results (Short-Time Fourier Transform), and applying FFT repeatedly within a sliding window over the data (Sliding Window Fourier Transform).

```{r, fig.show = 'hold', out.width='50%', fig.cap = "Periodogram from FFT (normalized for sample size and expressed in energy/amplitude) on simulated data with SNR=.446 (left) and with noise amplified to SNR=.05 (right). The estimates are poor at either noise level. The $f_\\frac{1}{12.5}$ (amplitude = .5) constituent is nearly buried in the higher noise."}
par(mar=c(4,4,.1,.1))
## this givs us close to the actual mags
plot(x = seq(0, .5, by = 1/8640),
     y=(2*abs(fft(simdat$x))/(nrow(simdat)))[1:4321],
     ylim = c(0,2), type = "l", ylab = "", xlab = "Frequency")
## scaled the original noise
plot(x = seq(0, .5, by = 1/8640),
     y = (abs(2*fft(simdat$x2))/(nrow(simdat)))[1:4321],
     ylim = c(0,2), type = "l", ylab = "", xlab = "Frequency")
```

### Waveform Shapes

Fourier analysis decomposes a function or signal into a weighted sum of sinusoidal components which form a complete basis set. This can be done even if the original signal is not generated by sinusoidal constituent because sine waves may be combined to approximate waveforms of arbitrary shapes. Typical examples include rectangular, triangular, and sawtooth patterns. The spectra in each case is characterized by energies at different multiples or *harmonics* of the fundamental frequencies at which these waveforms are present.

```{r, fig.show = 'hold', out.width = '50%', fig.cap = "Segment from a composite signal of 2 rectangular waveforms with periods .05 and .025, and its reconstruction (left); the energy density estimated by FFT (right). The reconstruction is performed using the 10 harmonics with highest energy and the $f_0$ component."}
par(mar=c(4,4,.1,.1))
simdat2 <- tibble(
  t = 0:(24*360),
  t_end = t+1,
  signal = c(rep(c(0,0,2,2,2,2,0,0,0,0,0,1,1,1,1,1,1,0,0,0,
                   0,0,2,2,2,2,0,0,0,0,0,0,0,0,0,0,0,0,0,0), 216),0))

sim2_fft <- fft(simdat2$signal)

filtered <- complex(8641, real = 0, imaginary = 0)
bins <- c(0,1,2,3,4,6,8,12,14,16,18)*216 + 1
filtered[c(bins, 8641 - bins[-1] + 2)] <- sim2_fft[c(bins, 8641 - bins[-1] + 2)]

recon_sim2_cplx <- fft(filtered, inverse = TRUE)
simdat2$recon <- Re(recon_sim2_cplx) / 8641

simdat2 %>%
  slice(1:80) %>%
  ggplot() +
  geom_step(aes(t, signal), color = "black") +
  geom_step(aes(t+.5, recon), color = "red") +
  labs(x = "t", y = "Energy")

plot(x = seq(0, .5, by = 1/8640),
     y=(2*abs(sim2_fft)/(nrow(simdat2)))[1:4321],
     ylim = c(0,2), type = "l", ylab = "Energy", xlab = "Frequency")
```



We return now to the temporal domain to introduce the KZ filter, upon which the KZFT method for spectral density estimation is based.

### Kolmogorov-Zurbenko Filter

The KZ filter is a sliding window arithmetic mean (simple average) applied *k*-times iteratively to a signal Let $\{X(t): t = 0, \pm1, \pm2,...\}$ be a real-valued, univariate time series at any measurement interval. For one iteration ($k=1$), the KZ filter applied at observation $X(t)$ is given by:
$$
KZ_{m,k=1}\big(X(t)\big) = \sum_{s=-(m-1)/2}^{(m-1)/2} X(t+s) \times \frac{1}{m}
$$
where $m$ odd is the bandwidth of the chosen window. This is simply the average of those values falling within the window centered at time $t$.^[Also known as a uniform kernel smoother in some fields.] With $k>1$ iterations the smoothing of the series is repeated on the result of the preceding iteration. The KZ filter can then be defined generally as
$$
KZ_{m,k}\big(X(t)\big) = \sum_{s=-k(m-1)/2}^{k(m-1)/2} X(t+s) \times a_{s}^{m,k}
$$
where coefficients $a_s^{m,k}$ are the observation weights determined by polynomial coefficients for the window size, given by
$$
a_s^{m,k} = \frac{c_s^{k,m}}{m^k} = \frac{1}{m^k}\sum_{r=0}^{k(m-1)}z^rc^{k,m}_{r-k(m-1)/2} = \frac{1}{m^k}(1+z+...+z^{m-1})^k
$$
The coefficients allow the iterative result to be calculated directly from the $k(m-1)+1$ values about $X(t)$ with increased efficiency. They can be interpreted as the probability mass function arising from a convolution of $k$ discrete uniform distributions on the interval $[X(t)-\frac{m-1}{2}, X(t)+\frac{m-1}{2}]$. The result for $k>1$ is a tapering window with finite support which approaches a gaussian distribution in the limit of $k$.

```{r uni_conv, fig.show='hold', fig.height=2.5, fig.width=4, fig.cap = "The filter produced by KZ(5,k) for k = 1,...,4."}
convuni <- tibble(t = c(1:19,1:19)-.5,
                  tend = c(1:19, 2:20)-.5) %>%
  mutate(d1 = c(rep(0, 8), coeff(5,1), rep(0, 6), rep(0, 7), coeff(5,1), rep(0, 7)),
         d2 = c(rep(0, 6), coeff(5,2), rep(0, 4), rep(0, 5), coeff(5,2), rep(0, 5)),
         d3 = c(rep(0, 4), coeff(5,3), rep(0, 2), rep(0, 3), coeff(5,3), rep(0, 3)),
         d4 = c(rep(0, 2), coeff(5,4), rep(0, 0), rep(0, 1), coeff(5,4), rep(0, 1))) %>%
  arrange(t, tend) %>%
  mutate(d1end = lead(d1, 1, 0),
         d2end = lead(d2, 1, 0),
         d3end = lead(d3, 1, 0),
         d4end = lead(d4, 1, 0))

grid.arrange(nrow = 2, heights = c(1,1.35), widths = c(1.175, 1),
convuni %>% 
  ggplot(aes(t, pmf)) +
  geom_segment(aes(x = t, xend = tend, y = d1, yend = d1end)) +
  annotate(geom = "text", label = "k=1", x = 17, y = .17) +
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank()), 
convuni %>% 
  ggplot(aes(t, pmf)) +
  geom_segment(aes(x = t, xend = tend, y = d2, yend = d2end)) +
  annotate(geom = "text", label = "k=2", x = 17, y = .17) +
  theme(axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank()),
convuni %>% 
  ggplot(aes(t, pmf)) +
  geom_segment(aes(x = t, xend = tend, y = d3, yend = d3end)) +
  scale_y_continuous(limits = c(0,.2)) +
  annotate(geom = "text", label = "k=3", x = 17, y = .17) +
  theme(axis.title.y = element_blank(),
        panel.grid = element_blank()),
convuni %>% 
  ggplot(aes(t, pmf)) +
  geom_segment(aes(x = t, xend = tend, y = d4, yend = d4end)) +
  scale_y_continuous(limits = c(0,.2)) +
  annotate(geom = "text", label = "k=4", x = 17, y = .17) +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank())
)
```

The KZ filter is classed as a "low-pass" filter in the field of signal analysis - high ("fast") frequencies are suppressed in the filter output as an effect of the nearby readings, while low ("slow") frequencies are allowed to pass to the output. A "cut-off" frequency can be determined using an arbitrarily chosen threshold level of power suppression; the cut-off frequency will be lower with increasing window size $m$ and iterations $k$. A formula for the commonly referenced "half-power point" is given approximately as
$$
\nu_c \approx \frac{\sqrt{6}}{\pi}\sqrt{\frac{1 - .5^{\frac{1}{2k}}}{m^2 - .5^{\frac{1}{2k}}}}
$$
The two parameters are often determined heuristically along with knowledge of the underlying process, the desired frequencies to be suppressed or retained, and the noise apparent in the signal. Working with hourly temperature data strongly governed by diurnal and annual constituents, for example, a 5-day window ($m=240$) and 1 or 2 iterations is sufficient to suppress the daily cycle, yet variance owing to exogenous factors (e.g. periods of cloud cover) and otherwise random fluctuations may dictate a larger window, perhaps 15 days, and maybe 2 or 3 iterations to realize the underlying annual component.

```{r kz_demo, fig.height = 3, fig.cap = "(left) Simulated signal, the $f_{168}$ constituent, and output of KZ filter applied with m=37 and k=1,2,3; (right) zoomed in and with the raw signal removed."}
kz_demo <- simdat %>%
               mutate(kz.37.1 = kza::kz(.$x, m = 37, k = 1),
                      kz.37.2 = kza::kz(.$x, m = 37, k = 2),
                      kz.37.3 = kza::kz(.$x, m = 37, k = 3)) %>%
               filter(time > 1800, time <= 2400) %>%
               select(time, c3, x, kz.37.1, kz.37.2, kz.37.3) %>%
               rename(signal = x, constituent3 = c3) %>%
               tidyr::pivot_longer(cols = c(signal, constituent3, kz.37.1, kz.37.2, kz.37.3),
                                   values_to = "energy", names_to = "name") %>%
               mutate(name = factor(name, levels = c("signal", "constituent3", "kz.37.1", "kz.37.2", "kz.37.3")))

grid.arrange(nrow = 1,
             kz_demo %>%
               ggplot(aes(time, energy, color = name)) +
               geom_line() +
               scale_color_manual(values = c("darkgrey", "black", "green", "purple", "orange")) +
               theme(legend.title = element_blank(),
                     legend.position = "inside",
                     legend.position.inside = c(.5, .93),
                     legend.background = element_blank()) +
               guides(color = guide_legend(nrow = 2, byrow = TRUE)),
             
             kz_demo %>%
               filter(time >= 1900, time <= 2100, 
                      name != "signal") %>%
               ggplot(aes(time, energy, color = name)) +
               geom_line() +
               scale_color_manual(values = c("red", "green", "purple", "orange")) +
               theme(legend.position = "none",
                     axis.title.y = element_blank())
)
```

The problem of determining optimal settings for these parameters may also be approached as an (unsupervised) machine learning task - i.e. by partitioning or re-sampling the data so that a grid search can be performed on the parameter space. Kernel smoothing methods generally do not provide forecasting ability, so some thought must be given to the sampling design. A loss function could be defined with terms representing the variances between smoothed results across re-samples or the out-of-fold/out-of-bag reconstruction error. For this simulated data, we find that $k=3$, $m=37$ (about 3 times the next highest constituent cycle-time) works well to obtain a result approximating the (known) lowest-frequency waveform.

```{r}
simdat %>%
  mutate(kz.71.1 = kza::kz(.$x, m = 37, k = 1),
         kz.71.2 = kza::kz(.$x, m = 37, k = 2),
         kz.71.3 = kza::kz(.$x, m = 37, k = 3),
         kz.71.4 = kza::kz(.$x, m = 37, k = 4),
         kz.71.5 = kza::kz(.$x, m = 37, k = 5)) %>%
  mutate(e1 = c3 - kz.71.1,
         e2 = c3 - kz.71.2,
         e3 = c3 - kz.71.3,
         e4 = c3 - kz.71.4,
         e5 = c3 - kz.71.5) %>%
  summarise(mse1 = mean(e1^2),
            mse2 = mean(e2^2),
            mse3 = mean(e3^2),
            mse4 = mean(e4^2),
            mse5 = mean(e5^2))
```

The KZ filter has useful properties which make it well-suited to de-noising signals (not limited to 1-dimensional types) and for separating waveforms embedded in noisy data. As a simple moving average it is robust to missing observations dispersed within the data (though longer runs of missingness become insurmountable as they approach and surpass the length of the window). At the most simple level it may be used for basic exploratory or graphical data analysis. For a more advanced application, it could be used as the basis of a filtering or convolution "layer" in neural network architectures to effectively extract embedded features for use downstream in the network.^[Zurbenko paper with image filtering from static] This paper's interest lies in its extension to the problem of estimating the spectral density of a signal, the focus to which we now turn.

## Part II - Kolmogorov-Zurbenko Fourier Transform

We discussed the challenge of estimating spectra in the presence of noise with FFT. We also reviewed the KZ filter as an iterated sliding window moving average which suppresses the variance from certain sources - namely, the variance attributable to constituent waveforms located above a threshold frequency, to non-cyclical features of the data generating process, and to the stochastic variance of e.g. a white noise. The Kolmogorov-Zurbenko Fourier Transform (KZFT) extends the KZ filter into the frequency domain as a $k$-times iterated Sliding Window Fourier Transform with window size $m$ applied at some frequency $\nu$. A symmetric filter applied in the frequency domain determines a *range* of frequencies whose energy may pass through the filter while suppressing those which are more distant. KZFT is classed then as a *band-pass* filter because it is used to smooth a power estimate at a given frequency by suppressing energy leakage from more distant frequencies thereby regularizing estimates. 

### Definition

KZFT is defined in Zurbenko and Porter (1998) as a "linear fourier filter" of the form
$$
F(t,\nu,M) = \frac{a}{2M+1}\sum^M_{s=-M} exp^{-is\nu}X(t+s)
$$
which is applied $K$ times iteratively to a signal of length $2MK + 1$. This can be expressed completely (with minor updates for consistency of notation and definitions) accounting for the iterations as:
$$
KZFT_{m,k,\nu_0}\big(X(t)\big) = F^{(k)}(t,\nu_0,m) = \sum^{k\frac{m-1}{2}}_{s={-k\frac{m-1}{2}}} a^{m,k}_s exp^{-i2s\nu_0}X(t+s)
$$
where $a^{m,k}_s$ are the same polynomial coefficient weights given previously. This provides an efficient estimator for power at the specified frequency $\nu_0$. Zurbenko (1986) shows that as an estimator for spectral density, KZFT is asymptotically nearly optimal from the point of view of mean square error under differentiability conditions on the spectral density.^[See Theorems 4.6 - 4.8, p.p. 96-97] To analyze the full spectrum within a window we apply the algorithm at every possible frequency up to $\frac{m-1}{2}$. Setting the window equal to the length of the original signal produces a power estimate similar to FFT in the first iteration. Subsequent iterations concentrate the weights locally and further attenuate the contribution of nearby frequencies. The 'sliding' of the window across the signal suggests that the power at a given frequency is estimated repeatedly - once at each index of the original signal - and the method produces power estimates which can then vary over time. KZFT is robust to non-stationarity as a result and can be used to detect changing patterns or power levels in any frequency.

We contrast the results of the classical periodogram produced by FFT to the KZ-periodogram obtained from applying a $KZFT_{4200,6}$ to our simulated signal with .05 SNR. Our signal is a stationary by design so the spectrum is estimated as the average result over the full signal. On visual inspection we see both that the estimated magnitudes for the three constituent frequencies are closer to the true values and the variance of estimates for unrelated frequencies is reduced.

```{r kzft_simdat, cache = TRUE}
kzp_sim_x2 <- kzp(simdat$x2, m = 4200, k = 6, ncores = 12)
saveRDS(kzp_sim_x2, "kzp_sim_x2.RDS")

# tibble(x = kzp_sim_x2$periodogram/2100,
#        f = (0:(length(kzp_sim_x2$periodogram)-1)/2100)) %>% View 
```

```{r, eval=FALSE}
kzp_sim_x2 <- readRDS("kzp_sim_x2.RDS")
```


```{r kzft_compare, fig.show = 'hold', out.width='50%', fig.cap = "Spectral density (energy) based on FFT (left) and on $KZFT_{\\nu,4200,6}$ (right) applied over all measurable frequencies within the window to produce the KZ-periodogram"}
par(mar=c(4,4,.1,.1))

plot(x = seq(0, .5, by = 1/8640),
     y = (abs(2*fft(simdat$x2))/(nrow(simdat)))[1:4321],
     ylim = c(0.03, 1.53), type = "l", ylab = "Energy", xlab = "Frequency")

plot(kzp_sim_x2, scale = "energy")
```

### Discussion

This definition or at least what it implies for implementation is worth a more detailed discussion. First, in contrast with DFT, KZFT does not utilize all values in the signal but only those falling within the bandwidth of a convolved uniform kernel centered at $X(t)$. Let us begin in the middle of the signal at $X(0)$. At the first iteration, the DFT calculated at index 0 will include the contribution of nearby values $X(t);\ t:= -\frac{m-1}{2},...,0,...,\frac{m-1}{2}$. The second iteration takes the result from the first over the same indices - so $m$ outputs from the first pass are needed. Then the original samples contributing in the second pass are $X(t);\ t:= -\frac{2(m-1)}{2},...,-\frac{m-1}{2},...,0,...,\frac{m-1}{2},...,\frac{2(m-1)}{2}$. For each output value $F^{(k)}(t)$, the $k(m-1)+1$ values centered about $t$ from the original signal contribute subject to the attenuation produced by the tapering window. Practically this means that we must choose from among several options how to handle cases where the window extends beyond the tails of the signal:

1. accept an output length reduced by $k(m-1)$ (half reduction at each end), in essence discarding these estimates;
1. pad the original series with zeroes so that the filter may be centered at each value of the input signal but suppressing (biasing toward 0) the power estimates near the tails;
1. allow for an asymmetrically shrinking window, down to a "half+1" window at the extremes (increasing the variance of estimates).

For a long series and/or short window this may have no practical consequence for estimates. Increasing either the window size or the number of iterations will of course increase the extent of the window loss at each end of the signal, which can meaningfully increase the variance of estimates when dealing with short (relative to the desired window length) time series.

KZFT is naturally robust to non-stationarity though this will also depend on the window size and time frame over which the non-stationarity is detectable. A window which is far larger than necessary if interested only in higher frequencies will permit a non-stationary component in the process to impact estimates if it is not nearly stationary within the window. In this case, a smaller window will resolve the issue while preserving the ability to detect the higher frequencies of interest. If non-cyclical trends or breaks/state changes are apparent within a more reasonable window then it may be necessary to de-trend or otherwise adjust the original signal prior to analysis, if possible.

Finally, available implementations are based on repeated direct calculations of the DFT - i.e. not utilizing the FFT algorithm. This allows for missing observations but at added computational cost. Within a window $m$ the algorithm is $O(m)$ complex at a single observation $X(t)$ for a single frequency $\nu_0$ and for one iteration. If we evaluate all frequency bins within the window (estimating the power spectrum) then it becomes $O(m^2)$ complex, equivalent to a DFT of a signal length $m$. Repeating at every value of the input signal (a sliding window) produces $O(nm^2)$ complexity *per iteration* and ultimately $O(knm^2)$ complexity in total. In other words, the computational cost scales poorly with window size. Consider a series of length 8,760 representing hourly readings taken over a year's time. Producing the KZ-periodogram with a window of length 169 (about a week) requires about 250.2 million operations per iteration - not terrible for modern hardware - but with a window length 4,369 (about 6 months) the overhead grows to 167.2 billion per iteration - or about 502 billion in total if $k=3$ iterations is desired. The prohibitive cost may demand some creativity by the user needing to analyze the lower end of a spectrum. Some options are:

1. Reduce the resolution of the signal by reducing the sampling frequency - i.e. reducing from an hourly sample rate to daily or weekly - which may be done either by aggregating the power (summing or averaging) of the original readings or by simply retaining every $S^*$'th sample and dropping those in between;
2. Parallelize the algorithm over multiple threads to reduce clock-time in proportion with the number of available cores;
3. If data is complete or missing values can be imputed beforehand then the algorithm can be implemented using FFT, providing improved performance of $O(k n m \log m)$.

Currently available implementations for computing the KZ-periodogram can produce impossibly long run-times for long signals and a large window. An implementation along the lines of option 3 above is sorely needed to expand the range of applications where KZFT can be practically applied. For now, a small modification can be made to the existing implementation of the `kza` R library to realize option 2, which can yield quite a large reduction in clock time (by a factor of $\frac{1}{\text{number of cores}}$) needed to evaluate a wider spectrum. A revised implementation is provided in Appendix 1.

## Part III: Application to Traffic Event Data

### Study Data

Our research question is whether the occurrence of traffic accidents around the capital region of New York in time is governed by cyclical components. The basis for our study is a dataset produced by Moosavi, et al (2019) compiling US Traffic Accident events broadcast by various entities including "US and state departments of transportation, law enforcement agencies, traffic cameras, and traffic sensors within the road-networks." These events are sourced from two traffic event API services, integrated (with likely duplicate events removed), geocoded, and published for research use.

```{r, tab.cap="Two source event records which overlap in time"}
cap %>%
  select(ID, Start_Time, End_Time, Zipcode, Description) %>%
  mutate(Description = substr(Description, 1, 30)) %>%
  filter(ID %in% c("A-2149438", "A-2149419")) %>%
  arrange(Start_Time) %>%
  kableExtra::kable(#format = "latex",
                    latex_options = "HOLD_position") %>%
  kableExtra::kable_styling(font_size = 10)
```

We subset the full data to those events occurring in selected capital region zip codes in 2019. There are `n_events` accident events recording in the time period for the area of study. Each event comes encoded with a start time and end time. We discretize these event records into hourly periods with the following logic:

1. Create a sequence of hourly periods spanning the time period covered by the research data;
1. "Overlap-left-join" the events to the hourly series based on `Start Time` and `End Time` of the event and for the hourly periods - each event is matched to the one or more periods it falls in;
1. Measure the portion of each hour (`(0, 1.0]`) which is covered by the event;
1. Summarize by hourly period the total duration of all events (any period with no events records 0 energy), denoting the resulting series as our signal `x(t)`.

```{r, tab.cap=" Discretization of the two events in Table 1 to create signal $X_t$"}
dat %>% 
  filter(t_start >= as.POSIXct('2019-04-05 06:00:00', tz = 'EST'),
         t_start < as.POSIXct('2019-04-05 11:00:00', tz = 'EST'))  %>%
  kableExtra::kable(#format = "latex",
                    latex_options = "HOLD_position") %>%
  kableExtra::kable_styling(font_size = 10)
```

An event lasting exactly one hour contributes 1.0 energy to the series in total; this contribution is fully allocated to the 12th hour of the given day if it begins at exactly 11:00:00 and ends at 12:00:00 EST. If the event begins at 11:15:00 and ends at 12:15:00 then 0.75 energy is allocated to hour 12 and 0.25 to hour 13. The theoretical range of `x(t)` is $[0,\infty)$, though at the high end this would be a very bad day for drivers in Manhattan. The maximum energy observed in any hour is `r round(max(dat$x), 4)`.

Figure 9 shows the smoothed empirical distribution of hourly energies, with and without the point mass at 0 (most hours having no events). Figure 10 displays a two-week period in the signal. We see here anecdotal evidence of at least two unsurprising patterns: energy is highest on weekdays and lowest on weekends; and energy appears to concentrate around the 8th and 17th hours.

```{r, fig.show='hold', fig.height = 3, fig.cap = "Smoothed distribution of measured hourly energy is shown in total (left) and for hours only with non-zero ernergy (right)."}
grid.arrange(nrow = 1,
             dat %>%
               ggplot(aes(x)) +
               stat_density(),
             dat %>%
               filter(x > 0) %>%
               ggplot(aes(x)) +
               stat_density() +
               theme(axis.title.y = element_blank())
             )
```

```{r, fig.show='hold', fig.height = 3.2, fig.cap="Signal energy for two-week period from Sunday, October 13 through Saturday, October 26, 2019"}
dat %>%
  filter(t_start >= as.POSIXct("2019-10-13 00:00:00", tz = 'EST'),
         t_start < as.POSIXct("2019-10-27 00:00:00", tz = 'EST')) %>%
  ggplot(aes(t_start, x)) +
  geom_line() +
  scale_x_datetime(breaks = c(seq.POSIXt(from = as.POSIXct("2019-10-13 08:00:00", tz = 'EST'),
                                         to = as.POSIXct("2019-10-26 08:00:00", tz = 'EST'),
                                         by = "24 hours"),
                              seq.POSIXt(from = as.POSIXct("2019-10-13 17:00:00", tz = 'EST'),
                                         to = as.POSIXct("2019-10-27 17:00:00", tz = 'EST'),
                                         by = "24 hours")),
                   date_labels = "%Y-%m-%d %H:%M") +
  scale_y_continuous(name = "Signal X(t)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 8),
        panel.grid.minor = element_blank())
```

```{r, eval = FALSE}
dat %<>%
  mutate(kz_7d.2 = kza::kz(x, m = 24, k = 2),
         kz_30d.5 = kza::kz(x, m = 24*30, k = 5)) %>%
  mutate(x_detrend = x * mean(x)/kz_30d.5,
         kz_7d.2.detrend = kza::kz(x_detrend, m = 24, k = 2),
         kz_30d.5.detrend = kza::kz(x_detrend, m = 24*30, k = 5))

dat %>%
  pivot_longer(c(kz_7d.2, kz_30d.5),
               names_to = "smooth", values_to = "value") %>%
  ggplot(aes(t_start, value, color = smooth)) +
  geom_line()

dat %>%
  pivot_longer(c(kz_7d.2.detrend, kz_30d.5.detrend),
               names_to = "smooth", values_to = "value") %>%
  ggplot(aes(t_start, value, color = smooth)) +
  geom_line()
```

### Analysis

We estimate the power spectral density of the signal to detect periodic components which may be embedded in it. In choosing appropriate parameters we anticipate waveforms with likely periods ranging from hours up to potentially monthly. We also recognize that traffic accidents are inherently random outcomes - "hourly traffic events" therefore presents a highly noisy signal and we will benefit from the smoothing produced by repeated iterations of KZFT. We select a window size $m=2184$ (13 weeks) and $k=3$ iterations given these considerations.

The periodogram shown in Figure 10 reveals four dominant constituents at frequencies .04167 (24 hours), .0833 (12 hours), .1250 (8 hours), and .005952381 (168 hours or 1 week), as well as a number of others (6 to 10) of intermediate to minor importance. The 10 most prominent frequencies are all higher-order harmonics of the fundamental weekly frequency, indicating the spectrum is largely approximating a non-sinusoidal weekly pattern. A number of other constituents (62.4, 19.675676, 12.066298 day cycles) are harmonics of a frequency with 364-day period which is not visible within the selected window. In other words, the signal is strongly determined by a weekly and possible annual pattern which are likely not best specified as a linear combination of sinusoids but can at least be approximated as such.

```{r kzp_big, cache=TRUE}
kzp_dat <- kzp(dat$x, m = 168*13, k = 3, ncores = 14)
saveRDS(kzp_dat, "kzp_dat_2184_3.RDS")
```

```{r}
## Constituents
kzp_results <- tibble(p = kzp_dat$power,
                      a = kzp_dat$energy) %>%
  mutate(f = (row_number()-1)/kzp_dat$window,
         period_hours = 1/f,
         period_days = period_hours/24) %>%
  arrange(desc(p)) %>%
  mutate(harmonic = round((f %% (1/168)), 8) == 0,
         include = (harmonic & row_number() <= 21))
```

```{r estimate_components, cache = TRUE}
f_include <- kzp_results %>%
  filter(include)

f_components <- matrix(nrow = nrow(dat), ncol = nrow(f_include))

for (i in 1:ncol(f_components)) {
  f_components[, i] <- kzft(dat$x, f = f_include$f[i], m = 168*13, k = 5)
}

saveRDS(f_components, "f_components.RDS")
```


```{r plot_kzp_dat, fig.show = 'hold', out.width='50%', fig.height = 5, fig.cap="(left) Power Spectral Density of hourly traffic accident signal; (right) scree plot of power spectrum constituents"}
par(mar=c(4,4,.1,.1))
plot.kzp(kzp_dat, scale = "energy", ylab = "Power (=Energy^2)")

kzp_results %>%
  arrange(desc(p)) %>%
  slice(-1) %>%
  mutate(rank = row_number()) %>%
  filter(rank < 800) %>%
  ggplot(aes(rank, p)) +
  geom_line() +
  geom_point(size = 1) +
  geom_segment(x = 21, xend = 21, y = -.0010, yend = .175, color = "red", linewidth = .2) +
  scale_y_continuous(name = "Power") +
  scale_x_continuous(name = "Rank")
```

#### Signal Reconstruction

Signal reconstruction is possible by summation of individual KZFTs estimated about the dominant constituent frequencies detected in the original signal. This is a typical application for filters and spectral analysis results when working with signals produced by physical phenomena - sounds, vibrations, ocean tides and such. In our context, we prefer to think of signal reconstruction in terms of producing an expectation or estimated risk function in time.

A scree plot provides visual indication of the importance of including each incremental constituent. Figure 12 shows the five dominant constituents individually over a week in March 2019 and the reconstructed signal using these components (plus the "DC" component `f0` which captures the constant or baseline power level) over the same two weeks. We observe in the reconstructed signal a pattern of twice-daily peaks occurring at 8:00 am and 5:00 pm, with higher peaks Monday-Friday compared to Saturday and Sunday. This demonstrates that a non-trivial calendar-weekly pattern can be approximated with just a small number of sinusoidal curves.

```{r, fig.show = 'hold', fig.height = 3, fig.width = 8, fig.cap = "(left) Top 5 constituent waveforms shown individually; (right) Event expectation based on these (includes $f_0$)"}
recon_5 <- data.frame(
  time = dat$t_start,
  f0 = 2*Re(f_components[,1]),
  f1 = 2*Re(f_components[,2]),
  f2 = 2*Re(f_components[,3]),
  f3 = 2*Re(f_components[,4]),
  f4 = 2*Re(f_components[,5]),
  f5 = 2*Re(f_components[,6]),
  recon = 2*Re(rowSums(f_components[,1:6])))

grid.arrange(nrow = 1,
             recon_5 %>%
               select(1, 3:7) %>%
               tidyr::pivot_longer(cols = c(f1:f5),
                                   names_to = "constituent",
                                   values_to = "power") %>%
               filter(time >= as.POSIXct('2019-06-10', tz = 'EST') - 5*60^2,
                      time < as.POSIXct('2019-06-17', tz = 'EST') - 5*60^2) %>%
               ggplot(aes(time + 5*60^2, power, color = constituent)) +
               geom_line() +
               scale_y_continuous(name = "Energy") +
               scale_x_datetime(breaks = seq.POSIXt(from = as.POSIXct("2019-06-11 00:00:00", tz = 'EST'),
                                                      to = as.POSIXct("2019-06-17 00:00:00", tz = 'EST'),
                                                      by = "24 hours"),
                   date_labels = "%Y-%m-%d %H:%M") +
               theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
                     axis.title.x = element_blank(),
                     legend.position = "none"),
             
             recon_5 %>%
               filter(time >= as.POSIXct('2019-06-10', tz = 'EST') - 5*60^2,
                      time < as.POSIXct('2019-06-17', tz = 'EST') - 5*60^2) %>%
               ggplot(aes(time+5*60^2, recon)) +
               geom_line() +
               scale_y_continuous(name = "Expected Events") +
               scale_x_datetime(breaks = c(seq.POSIXt(from = as.POSIXct("2019-06-10 08:00:00", tz = 'EST'),
                                                      to = as.POSIXct("2019-06-17 08:00:00", tz = 'EST'),
                                                      by = "24 hours"),
                                           seq.POSIXt(from = as.POSIXct("2019-06-10 17:00:00", tz = 'EST'),
                                                      to = as.POSIXct("2019-06-17 17:00:00", tz = 'EST'),
                                                      by = "24 hours")),
                   date_labels = "%Y-%m-%d %H:%M") +
               theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        axis.title.x = element_blank())
)
```

From the estimated PSD and the scree plot it appears we can complicate the signal reconstruction and explain greater systematic variation by including more than 5 components. In fact, the lower estimator variance of KZFT has reduced the risk of "over-fitting" noise in the original signal, suppressing estimates for unrelated frequencies toward 0 energy. We select the top 15 frequencies by rank which are also harmonic constituents of the fundamental weekly frequency. In pure rank terms these all fall within the top 20 frequencies ranked by power estimate.

The resulting expected risk function reveals further interesting and intuitive temporal patterns. Weekdays exhibit the twice daily peaks at 08:00 and 17:00 mentioned earlier, but we now observe a single peak on Saturdays and Sundays occurring about mid-day, between the 12:00 and 13:00 hours. We also see smaller spikes in event incidence emerged in the late evening/early morning hours of Fridays and Saturdays, and less intuitively, Wednesdays. This can indicate increased roadway traffic in those hours compared to weeknights, higher incidence of impaired drivers, poor approximation of the true pattern, or some combination of these Finally, inclusion of the additional waveforms has increased the variance of the reconstructed signal, which now exhibits higher peak levels and faster ramping to and from these peaks.

```{r, fig.show = 'hold', fig.height = 4, fig.cap="(top) Signal reconstruction based on 15 constituent frequencies and $f_0$; (bottom) Average hourly event for 2019"}
grid.arrange(nrow = 2,
             data.frame(
               time = dat$t_start,
               recon = 2*Re(rowSums(f_components))) %>%
               filter(time >= as.POSIXct('2019-06-03', tz = 'EST') - 5*60^2,
                      time < as.POSIXct('2019-06-10', tz = 'EST') - 5*60^2) %>%
               ggplot(aes(time+5*60^2, recon)) +
               geom_line() +
               scale_x_datetime(breaks = c(seq.POSIXt(from = as.POSIXct("2019-06-03 08:00:00", tz = 'EST'),
                                                      to = as.POSIXct("2019-06-17 08:00:00", tz = 'EST'),
                                                      by = "24 hours"),
                                           seq.POSIXt(from = as.POSIXct("2019-06-03 17:00:00", tz = 'EST'),
                                                      to = as.POSIXct("2019-06-17 17:00:00", tz = 'EST'),
                                                      by = "24 hours")),
                                date_labels = "%Y-%m-%d %H:%M",
                                expand = c(0,0)) +
               theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 8),
                     axis.title.x = element_blank(),
                     panel.grid.minor = element_blank()),
             
             dat %>%
               mutate(Year = year(t_start) %>% as.factor,
                      Day = lubridate::wday(t_start, label = TRUE) %>%
                        factor(levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")),
                      Hour = hour(t_start)) %>%
               group_by(Year, Day, Hour) %>%
               summarise(x = sum(x)) %>%
               ungroup() %>%
               ggplot(aes(x = Hour, y = x)) +
               geom_bar(stat = "identity", position = position_dodge(), width = 1) +
               theme(panel.spacing.x = unit(0, "lines"),
                     panel.grid.minor = element_blank()) +
               scale_x_continuous(breaks = c(0, 8, 17),
                                  expand = c(0,0),
                                  name = "Day of Week - Hour") +
               scale_y_continuous(labels = scales::label_number(accuracy = .1)) +
               facet_grid(~ Day)
)
```

Because the signal is strongly determined by the fundamental weekly frequency and its harmonics, we are justified in ultimately selecting some method based in the time domain to derive final estimates. We may compare the results above to a naive estimator for day-hour periods - i.e. fitting a regression to these contrasts or simply calculating the mean by group. We note though that such a model would utilize 168 ($24 \times 7$) degrees of freedom, ignoring order and making no assumptions about smoothness in time.^[Our signal can be split into about 4 independent blocks, though we do not treat them independently. It would seem that $4 \times 2 \times 15$ would serve as a good estimate for total d.f. for a reconstruction based on 15 constituents derived from a single iteration of KZFT, but this would decrease with additional iterations - Formula 1.1.19 Loneck?]

<!-- ### Analysis 2 -->

<!-- Non-stationariy over multiple years (time allowing) -->

```{r, eval = FALSE}
data.frame(
  time = dat$t_start,
  recon = 2*Re(f_components[, 1])) %>%
  ggplot(aes(time, recon)) +
  geom_line()
```

### Discussion

etc

\newpage
\theendnotes

\newpage
## Bibliography

\begin{hangparas}{.25in}{1}

Dominguez, A. "Highlights in the History of the Fourier Transform." \emph{IEEE Pulse}, Volume 7, Issue 1, January 2016. https://ieeexplore.ieee.org/stamp/stamp.jsp?arnumber=7389485.

Zurbenko I. \emph{The Spectral Analysis of Time Series}. New York: North-Holland; 1986.

\end{hangparas}

\newpage
## Appendix 1: parallelized `kzp()` (Unix/Linux/MacOS)

We provide a modified implementation of the `kzp()` method from library `kza`. This version provides parallelization over the frequency bins - i.e. the energy estimates at each frequency within the window are produced concurrently across multiple cores and combined in the result. This can provide a significant reduction in processing time (see additional notes below). The `parallel` package is required. An implementation for Windows is provided in Appendix 2.

The method has also been modified to optionally produce estimates for a smaller range of frequencies within a window, specified in the `f_range` argument. This preserves frequency resolution without requiring processing of every frequency bin. The user may supply a length 2 vector specifying the minimum and maximum frequency bin to evaluate; `f_range=c(0,m-1)` is the default and original behavior.

Lastly, the method has been modified to return the estimated energy and power normalized for window size `m` and iterations `k`, in addition to the original, un-normalized result. Each is accessible in the return object and selectable for plotting by `plot.kzp()` (see Appendix 3).

You can determine the number of cores available on your machine prior to calling the method. It is advisable to always use 1 or 2 fewer than total available cores, so that resources will be left for other processes and the system remains responsive.

```{r, eval = FALSE, echo = TRUE}
## determine how many cores available and how many to use
library(parallel)
n_cores <- detectCores() - 2
```

The user allocates the desired number of cores in the method call (and optionally may also supply a frequency range).

```{r, eval = FALSE, echo = TRUE}
kzp <- function(y, m=length(y), k=1, f_range = NULL, ncores = n_cores)
{
  M<-(m-1)*k+1
  n=length(y)
  
  ## default to full spectrum as determined by m
  ## but allow user-specified range (provided in frequency-bin)
  if (is.null(f_range)) {
    f_est <- c(0,m-1)
  } else {
    f_est <- f_range
  }
  
  ## parallelized over the frequencies to be evaluate
  z_tmp <- parallel::mclapply(
    X = f_est[1]:f_est[2],
    FUN = function(i, x, m, k) kzft(x,f=i/m,m,k=k),
    x = y, m = m, k=k,
    mc.cores = ncores
  )
  
  z <- matrix(unlist(z_tmp), nrow = n, ncol = f_est[2]-f_est[1]+1)
  
  d <- apply(z, 2, function(z) {(abs(z)^2)*M})
  a <- colMeans(d, na.rm=TRUE)
  
  if(is.null(f_range)) {
    a <- a[1:(m/2)]
    f_range <- c(0, length(a)-1)
  } else {
    a <- a[1:(f_range[2] - f_range[1] + 1)]
  }
  b <- 2*sqrt(a/M)
  c <- b^2
  
  structure(list(
    periodogram = a, ## original
    energy = b, ## estimated energy
    power = c, ## estimated power
    window=m,
    f_range=f_range, ## for user-specified frequency ranges
    k=k,
    var=var(y),
    smooth_periodogram=NULL,
    smooth_method=NULL,
    call=match.call()
  ),
  class = "kzp")
}
```

This implementation is based on `parallel::mclapply()` which relies on forking and therefore only works on Unix-like systems (Linux, Apple OS). Unlike Unix systems, Windows OS do not support POSIX system forking, an efficient method which allows child processes to access objects in shared memory. The method will not throw an error if called on Windows but will simply process with a single core - the Windows user must be aware they will not obtain the desired parallelization and will not be alerted to this.

Concurrently evaluating frequencies in the spectrum of a window can provide significant reduction of processing time (wall clock) compared to the sequential, single-threaded implementation currently available. The reduction in run time is in proportion to the number of CPU cores allocated; e.g. allocating 10 cores (available on many higher-end laptops) will reduce the run time by 90%. A more powerful server may contain upwards of 300 cores, allowing an extreme reduction of processing time (if you can grab them all without annoying your neighbors). This is critical to enabling analysis of long signals (i.e. many thousands of readings) with a large window. Tidal and atmospheric data, for example, are available today at 6-minute resolution going back over a decade; the lunar nodal cycle is ~18.6 years long.

\newpage

## Appendix 2: parallelized `kzp()` (Windows)

A similar implementation can be realized with `parallel::parLapply()` which relies on PSOCK clusters supported by Windows. A cluster of child processes is created and objects in memory are copied to each, making it less efficient than forking. The user must be mindful of their total system memory and the size of required data objects (i.e. the signal object) at the risk of hanging their R session or crashing the system. Only distribute objects needed by `kzft()`.

```{r, eval = FALSE, echo = TRUE}
cl <- makeCluster(n_cores, type = "PSOCK")
clusterExport(cl, varlist = c("signal_object", "kzft"))
```

The implementation of `kzp()` provided above would be re-written slightly replacing the command at line 15 with:

```{r, eval = FALSE, echo = TRUE}
  z_tmp <- parallel::parLapply(
    cl = cl,
    X = f_est[1]:f_est[2],
    FUN = function(i, x, m, k) kzft(x,f=i/m,m,k=k),
    x = y, m = m, k=k
  )
```

After calling `kzp()`, you must stop the cluster to free the allocated cores for regular system use.

```{r, eval = FALSE, echo = TRUE}
stopCluster(cl)
```

\newpage
## Appendix 3: Enhanced `plot.kzp()`

The `plot.kzp()` method from `kza` is rewritten to accommodate several options for scale of the plotted spectral density with the `scale` argument. The original option is preserved - this is something like total power but increases with `k`. Options for the energy and power spectral densities are added. We have also added the option to include or exclude $f_0$ (the DC component) from the plot - including it will often conceal what is happening across the spectrum due to scale differences.

Support for plotting smoothed spectral densities is removed out of convenience rather than necessity.

```{r, eval = FALSE, echo = TRUE}
plot.kzp <- function(x, remove0 = TRUE, scale = NA, ylab = '', ...) {
  if (!is.na(scale)) {
    stopifnot("scale must be one of 'original', 'power', or 'energy'"
                = scale %in% c("original", "power", "energy")) 
  }
  
  if(is.na(scale) | scale == "original") {
    p <- x$periodogram
  } else if (scale == "energy") {
    p <- x$energy
  } else if (scale == "power") {
    p <- x$power
  }
  
  if(x$f_range[1] == 0 & remove0) {
    dz <- p[-1]
    omega <- ((x$f_range[1]+1):x$f_range[2])/x$window
    plot(omega, dz, type="l", xlab="Frequency", ylab=ylab)
  } else  {
    dz <- p
    omega <- (x$f_range[1]:x$f_range[2])/x$window
    plot(omega, dz, type="l", xlab="Frequency", ylab=ylab)
  }
}
```

\newpage
## Appendix 4: `kzft()`

The `kzft()` method as provided in R library `kza` (with some minor formatting changes) is provided here for completeness and posterity. The package is no longer maintained and will be archived by CRAN once it begins to fail automated build checks. Note that this method is not exposed upon loading the library as is expected of R packages; the user must source the method into their environment from file `kzft.R` which is bundled in the package or can be obtained from the CRAN GitHub mirror.

```{r, eval = FALSE, echo = TRUE}
kzft <- function(x, f=0, m=1, k=5)
{
  h <- floor((m-1)/2)
  t  <- ceiling((m-1)/2)
  x <- append(x, rep(NA,t*k), after = length(x) )
  x <- append(x, rep(NA,h*k), after = 0)
  z <- as.complex(x)
  
  for(i in 1:k) {
    s     <- rep(0, length(z))
    count <- rep(0, length(z))
    count <- count + !is.na(z)
    z[is.na(z)] <- 0
    s <- as.complex(s) + z
    
    for(j in 1:h) {
      # add offset
      z   <- c(rep(NA, j), x[1:(length(x)-j)])
      z <- z * exp(complex(real = 0, imaginary = 2*pi*f*j))
      count <- count + !is.na(z)
      z[is.na(z)] <- 0
      s <- s + z
    }
    for(j in 1:t) {
      z   <- c(x[(j+1):length(x)], rep(NA, j))
      z <- z * exp(complex(real = 0, imaginary = -2*pi*f*j))
      count <- count + !is.na(z)
      z[is.na(z)] <- 0
      s <- s + z
    }
    z <- s/count
    x<-z
  }
  z <- z[(t*k+1):(length(z)-(h*k))]
  return(z);
}
```

```{r}
knitr::knit_exit()
```
